---
title: Mapping polling errors in US elections, 2016-20
author: ''
date: '2022-11-06'
slug: polling-errors-map
categories:
  - Politics
  - Data science
  - Academic
tags:
  - statistics
  - politics
  - polling
  - survey research
draft: no
summary: "Some folks like to look at current polls and poll averages and make a mental adjustment to the results and see how they'd look if the polls missed the way they did in the recent past. But I think even for people who follow
this stuff pretty closely, it can be hard to remember all those details. My goal here is to help with that kind of thing while also showing that the narrative about polling errors is probably an overgeneralization."
image:
  caption: ''
  focal_point: ''
photoswipe: false
disable_jquery: true
math: false
mathjax: false
output:
  blogdown::html_page:
    toc: true
---

<style type="text/css" media="screen">

.article-container {
  max-width:1000px !important;
}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warn = FALSE)
```

If you follow US politics at all, you've probably heard a few things about 
polling in the past few election cycles:

* They have been "wrong"
* They have been biased towards Democrats
* They have been worse in some places than others

And there's some truth to all of this! But I think even for people who follow
this stuff pretty closely, it can be hard to remember all the details. Some 
folks like to look at current polls and poll averages and make a mental 
adjustment to the results and see how they'd look if the polls missed the way 
they did in the recent past. My goal here is to help with that kind of thing.

Below are three types of maps:

1. Average polling error for each state in the past 3 elections
2. Average polling bias for each in the past 3 elections
3. The 2022 Senate election polling averages 

Some states seem to be easier to poll than others. The **first map** is meant to 
highlight the ones where polls are generally more on the mark, regardless of 
whether they tilt towards one or the other party. It goes without saying that
high levels of bias also cause error, but some places may have
little bias but the typical poll still misses by a good deal.

The **second map** shows polling bias instead. Here we are looking at how much
the average polling margin tends to miss towards one party or the other compared
to the actual results. I've shaded states blue for pro-Democrat bias (meaning 
the polling average *overestimates* the Democrats) and red for pro-Republican 
bias. Lighter shades mean the bias is closer to zero.

Finally, the **third map** lets you take the current (as of 11/6)
FiveThirtyEight polling averages, which is shown in the default view, and see
how the margins would change if you applied the levels of bias seen in that 
state in past elections. Let's just say the Democrats will be rooting for a 
2018 miss and the Republicans will probably be happy with anything else.

The polling data used to calculate bias and error is for President, Senate, and
Governor only. No other statewide races, ballot issues, or Congressional races 
are used here due to data availability and other issues.

**Hover over states (or tap on mobile) to see the exact number**.

### Important note about methods

These numbers are not derived from simply averaging all public polls and 
comparing them to the results. Instead, I'm using a statistical model to
adjust for the quality and types of polls conducted in each race. So the
estimates are ultimately quite similar to what you'd get if you compared 
the projected vote share from [FiveThirtyEight](https://fivethirtyeight.com) 
(which is also based on polls with adjustments for pollster quality, 
partisanship, and some other things) to the actual votes.

See the "[How this works](#how-this-works)" section for more info.

```{r include = FALSE}
library(tidyverse)
library(readr)
library(lubridate)
library(plotly)
library(brms)
library(rjson)
library(scales)
library(jtools)
library(magrittr)

usgeo <- fromJSON(file = "https://eric.clst.org/assets/wiki/uploads/Stuff/gz_2010_us_040_00_500k.json")

d <- read_csv("./polling-error-files/raw-polls.csv") %>%
  mutate(
    # statewide = stringr::str_detect(race, "Sen\\-G|Gov\\-G|Pres\\-G"),
    statewide = type_simple %in% c("Sen-G", "Gov-G", "Pres-G"),
    partisan = case_when(
      is.na(partisan) ~ "No",
      TRUE ~ partisan
    ),
    year = case_when(
      year == 2017 ~ 2018,
      year == 2019 ~ 2020,
      TRUE ~ year
    ),
    partisan = relevel(factor(partisan), ref = "No"),
    days_till_election = as.numeric(mdy(electiondate) - mdy(polldate)),
    yearf = factor(year),
    methodology = factor(methodology),
    methodology = relevel(methodology, ref = "Live Phone"),
    weight = 0.5^(as.numeric(days_till_election)/30) * sqrt(min(c(samplesize, 2000))/600)
  ) %>%
  filter(year >= 2016, statewide, location %nin% "PR") %>%
  mutate(year = factor(year))

options(Ncpus = 6, mc.cores = 6)
library(brms)

# fit <- brm(mvbf(bias ~ year + partisan + log(samplesize) + days_till_election +
#            (year | pollster_rating_id) + (year | location/type_simple),
#            error ~ year + partisan + log(samplesize) + days_till_election +
#            (year | pollster_rating_id) + (year | location/type_simple)),
#            data = d, iter = 4000)
# saveRDS(fit, "fitcombined.RDS")
fit <- readRDS("./polling-error-files/fitcombined.RDS")

state_biases <- make_predictions(fit, "year", 
                                 at = list(location = unique(d$location), 
                                           days_till_election = 0, 
                                           samplesize = 800), 
                                 re.form = ~ (year | location), resp = "bias")

state_biases %<>%
  group_by(location) %>%
  mutate(
    bias16 = case_when(
      year == 2016 ~ bias,
      TRUE ~ NA_real_
    ),
    bias18 = case_when(
      year == 2018 ~ bias,
      TRUE ~ NA_real_
    ),
    bias20 = case_when(
      year == 2020 ~ bias,
      TRUE ~ NA_real_
    )
  ) %>%
  summarize(
    bias16 = mean(bias16, na.rm = T),
    bias18 = mean(bias18, na.rm = T),
    bias20 = mean(bias20, na.rm = T)
  )

dgyears <- d %>%

  group_by(location, year) %>%
  summarize(
    bias_u = mean(bias, na.rm = T),
    bias = weighted.mean(bias, w = weight, na.rm = T),
    polls = n(),
    races = length(unique(race_id))
  ) 

dgyears %<>% 
  ungroup() %>%
  mutate(
    races_2020 = case_when(
      year == 2020 ~ races
    ),
    races_2018 = case_when(
      year == 2018 ~ races
    ),
    races_2016 = case_when(
      year == 2016 ~ races
    ),
    polls_2020 = case_when(
      year == 2020 ~ polls
    ),
    polls_2018 = case_when(
      year == 2018 ~ polls
    ),
    polls_2016 = case_when(
      year == 2016 ~ polls
    )
  ) %>% 
  group_by(location) %>%
  summarize(
    races_2020 = mean(races_2020, na.rm = T),
    races_2018 = mean(races_2018, na.rm = T),
    races_2016 = mean(races_2016, na.rm = T),
    polls_2020 = mean(polls_2020, na.rm = T),
    polls_2018 = mean(polls_2018, na.rm = T),
    polls_2016 = mean(polls_2016, na.rm = T),
    polls = sum(polls_2016, polls_2018, polls_2020, na.rm = T)
  ) %>%
  mutate(
    races_2018 = ifelse(is.na(races_2018), yes = 0, no = races_2018),
    polls_2018 = ifelse(is.na(polls_2018), yes = 0, no = polls_2018),
  )


  
state_biases <- left_join(state_biases, dgyears)
state_biases$bias_avg <- sapply(1:nrow(state_biases), function (x) {
  weighted.mean(state_biases[x, c("bias16", "bias18", "bias20")], w = state_biases[x, c("polls_2016", "polls_2018", "polls_2020")])
})

states_names_abbs <- bind_cols(abb = state.abb, name = state.name)

dg <- left_join(state_biases, states_names_abbs, by = c("location" = "abb"))


### Errors
state_errors <- make_predictions(fit, "year", 
                                 at = list(location = unique(d$location), 
                                           days_till_election = 0, 
                                           samplesize = 800), 
                                 re.form = ~ (year | location), resp = "error")

state_errors %<>%
  group_by(location) %>%
  mutate(
    error16 = case_when(
      year == 2016 ~ error,
      TRUE ~ NA_real_
    ),
    error18 = case_when(
      year == 2018 ~ error,
      TRUE ~ NA_real_
    ),
    error20 = case_when(
      year == 2020 ~ error,
      TRUE ~ NA_real_
    )
  ) %>%
  summarize(
    error16 = mean(error16, na.rm = T),
    error18 = mean(error18, na.rm = T),
    error20 = mean(error20, na.rm = T)
  )
  
state_errors <- left_join(state_errors, dgyears)
state_errors$error_avg <- sapply(1:nrow(state_errors), function (x) {
  weighted.mean(state_errors[x, c("error16", "error18", "error20")], w = state_errors[x, c("polls_2016", "polls_2018", "polls_2020")])
})

dge <- left_join(state_errors, states_names_abbs, by = c("location" = "abb"))
```

## Polling Error 

```{r out.width = "100%", out.height = "600px"}
fig <-  plot_ly(dge) %>% add_trace(
  type = "choropleth", locations = ~name, featureidkey = "properties.NAME",
  geojson = usgeo,  z = ~error_avg, color = ~error_avg, 
  colors = "Purples",
  zmin = 2, zmax = 8,  hovertemplate = '%{z:2d}%<extra>%{location}</extra>',
  colorbar = list(title = list(text = "Average Error", side = "top"),
                  len = 150, lenmode = "pixels",
                  orientation = "h", y = 0,
                  tickformat = "2d%%",
                  tick0 = 0,
                  tickvals = c(2, 5, 8),
                  ticktext = c("2%", "5%", "8%"))
) %>%
  layout(
    geo = list(
      scope = 'usa'
    ),
    dragmode = FALSE,
    updatemenus = list(
      list(
        type = "buttons",
        xanchor = 'center',
        yanchor = "top",
        direction = "right",
        pad = list('r'= 0, 't'= 10, 'b' = 10),
        x = 0.5,
        y = 1.17,
        buttons = list(
          list(method = "restyle",  
               args = list(list("z" = list(~error_avg), "color" = list(~error_avg))), 
               label = "2016-20 average"),
          list(method = "restyle",
               args = list(list("z" = list(~error16), "color" = list(~error16))),  
               label = "2016"),
          list(method = "restyle", 
               args = list(list("z" = list(~error18), "color" = list(~error18))),
               label = "2018"),
          list(method = "restyle",
               args = list(list("z" = list(~error20), "color" = list(~error20))),
               label = "2020")))  
    )
  )

fig
```


## Polling Bias

```{r out.width = "100%", out.height = "600px"}
fig <-  plot_ly(dg) %>% add_trace(
  type = "choropleth", locations = ~name, featureidkey = "properties.NAME",
  geojson = usgeo,  z = ~bias_avg, color = ~bias_avg, 
  colorscale = list(c(0, 0.5, 1), c(muted("red"), "white", muted("blue"))),
  zmin = -8, zmax = 8,  hovertemplate = 'D %{z:+2.1d}%<extra>%{location}</extra>',
  colorbar = list(title = list(text = "Average Bias", side = "top"),
                  len = 150, lenmode = "pixels",
                  orientation = "h", y = 0,
                  tickformat = "D%+2d%%",
                  tick0 = 0,
                  tickvals = c(-8, -4, 0, 4, 8),
                  ticktext = c("D -8%", "", "No bias", "", "D +8%"))
) %>%
  layout(
    geo = list(
      scope = 'usa'
    ),
    dragmode = FALSE,
    updatemenus = list(
      list(
        type = "buttons",
        xanchor = 'center',
        yanchor = "top",
        direction = "right",
        pad = list('r'= 0, 't'= 10, 'b' = 10),
        x = 0.5,
        y = 1.17,
        buttons = list(
          list(method = "restyle",    
               args = list(list("z" = list(~bias_avg), "color" = list(~bias_avg))),    
               label = "2016-20 average"),
          list(method = "restyle",   
               args = list(list("z" = list(~bias16), "color" = list(~bias16))),   
               label = "2016"),
          list(method = "restyle",    
               args = list(list("z" = list(~bias18), "color" = list(~bias18))),  
               label = "2018"),
          list(method = "restyle", 
               args = list(list("z" = list(~bias20), "color" = list(~bias20))), 
               label = "2020")))
    )
  )

fig
```

## 2022 Senate polling averages with bias adjustments 

```{r out.width = "100%", out.height = "600px"}
projs <- read_csv("./polling-error-files/senate_steps_2022.csv") %>%
  filter(displaystep == 1) %>%
  # select(district, margin) %>%
  mutate(
    state = str_extract(district, "..")
  )

projs <- full_join(projs, dg, by = c("state" = "location"))
projs %<>%
  mutate(
    margin_avg = margin - bias_avg,
    margin16 = margin - bias16,
    margin18 = margin - bias18,
    margin20 = margin - bias20
  )
projs <- filter(projs, district != "OK-S2", !is.na(margin))
 
fig <-  plot_ly(projs) %>% add_trace(
  type = "choropleth", locations = ~name, featureidkey = "properties.NAME",
  geojson = usgeo,  z = ~margin, color = ~margin, 
  colorscale = list(c(0, 0.5, 1), c(muted("red"), "white", muted("blue"))),
  zmin = -10, zmax = 10,  hovertemplate = 'D %{z:+2.1d}%<extra>%{location}</extra>',
  colorbar = list(title = list(text = "Margin of victory", side = "top"),
                  len = 150, lenmode = "pixels",
                  orientation = "h", y = 0,
                  tickformat = "D%+2d%%",
                  tick0 = 0,
                  tickvals = c(-10, -5, 0, 5, 10),
                  ticktext = c("D -10%", "", "Even", "", "D +10%"))
) %>%
  layout(
    geo = list(
      scope = 'usa'
    ),
    dragmode = FALSE,
    updatemenus = list(
      list(
        type = "buttons",
        xanchor = 'center',
        yanchor = "top",
        direction = "right",
        pad = list('r'= 0, 't'= 10, 'b' = 10),
        x = 0.5,
        y = 1.17,
        buttons = list(
          list(method = "restyle",    
               args = list(list("z" = list(~margin), "color" = list(~margin))),    
               label = "Current polling average"),
          list(method = "restyle",    
               args = list(list("z" = list(~margin_avg), "color" = list(~margin_avg))),    
               label = "with 2016-20 average bias"),
          list(method = "restyle",   
               args = list(list("z" = list(~margin16), "color" = list(~margin16))),   
               label = "with 2016 bias"),
          list(method = "restyle",    
               args = list(list("z" = list(~margin18), "color" = list(~margin18))),  
               label = "with 2018 bias"),
          list(method = "restyle", 
               args = list(list("z" = list(~margin20), "color" = list(~margin20))), 
               label = "with 2020 bias")))
    )
  )

fig
```

## How this works

As already mentioned, this is not a simple polling average in the sense of 
finding the mean of all polls and reporting just that. My opinion was that this
just wasn't going to be all that informative since the number, quality, and type
of polls would vary too much from place to place and time to time. In other 
words, I basically agree with the the FiveThirtyEight approach to poll averaging.

To generate the numbers you see in the maps, I created a statistical model to 
generate average polling error and bias estimates for each state in each year
(off-year elections in 2017 and 2019 were classified as 2018 and 2020). I'll 
describe that in more detail the statistically inclined at the end of this 
section. Basically, what this does for us is applies adjustments for the 
quality/tendencies of the individual pollster, the partisanship of the pollster
(if any), the sample size, and 
the time until the election when the poll was fielded. This means if one state
seems to attract low-quality partisan pollsters, it won't look as bad in my 
model as it would in a simple average. 

The best way to think about it is the cycle-specific numbers 
represent my best estimate of **the polling error/bias you would expect from 
an average quality non-partisan pollster using a sample size of about 800 on
the day before election day**. In some states, that's pretty similar to the 
literal polling average. In some others, it's rather different due to usually
fewer polls from a subset of pollsters who do not represent the industry well 
(either because they do better or worse). 

To create the 3 year average, I simply take the 3 cycle-specific estimates and  
calculate a weighted mean. Why weighted? I weight by the number of polls per 
cycle. I could do it in a more sophisticated way, but it didn't ordinarily 
change the results. I could also have fit a separate model to create this 
estimate, but it would have given something quite similar.

### Data

The **data** comes from 
[FiveThirtyEight's public data](https://github.com/fivethirtyeight/data/tree/master/pollster-ratings) 
used for generating their
pollster grades. I make no use of their pollster grades, but they conveniently 
include all late-cycle polls (i.e., within 3 weeks of election day) along with
pre-calculated error and bias for each poll (so I don't have to look up the 
results for every single race).

I do not use the data from primaries because they cannot have a partisan bias
and generally have so much more polling error that it would swamp everything 
else. Combine that with the greatly differing methods for doing primaries 
(e.g., Iowa's caucus system) and it just didn't seem right to include even if 
just for the polling error part.

I did not include statewide races for the House, like in Montana, Wyoming, and
Alaska. I may change that in the future. I also do not include Presidential 
results for the congressional districts in Nebraska and Maine that cast their 
votes according to their district vote rather than the statewide vote. I'm open 
to it, but can't find a shapefile that has the 50 states *and* cutouts for those
districts. 😅 

### The model

I'm a one-trick pony, so whether it's (calculating the 
value of field goal kickers)[/post/evaluating-kickers] or mapping polling 
errors, I'm reaching for a multilevel model. There are some serious advantages
in this case. The data are grouped in a few ways:

* By state
* By pollster 
* By election type (President/Senate/Governor)
* By year of election

Now there aren't actually enough separate election years to accomplish much
with treating it as a leveling factor in the multilevel model, but the others
do have some value. This is especially true because many of the pollsters and 
states won't contribute much data so we need a principled way to use their data
without overfitting. And multilevel models let me make adjustments for other
things like pollster partisanship, time until the election, and sample size. 
(Note: I considered adding adjustments by method but there are *so many* thanks 
to pollsters combining many at once. It was a real mess and not adding much.)

With so many "groups," many of which have little data associated, and a desire 
to produce useful preditions, the obvious option (for me, a one-trick pony) was
Bayesian estimation via [`brms`](https://paul-buerkner.github.io/brms/). I love
this R package. With just a few lines of code I was able to fit a relatively
complex model that includes multiple dependent variables (error and bias) that
allows the residuals of those two dependent variables to correlate. With a 
few more lines of code, I could generate state- and year-level predictions via
some of the conveniences in my [`jtools` package](https://jtools.jacob-long.com).

Seriously, see below how brief the most intellectually challenging steps are.

Model fit:

```R
brm(mvbf(
  bias ~ year + partisan + log(samplesize) + days_till_election +
  (year | pollster_rating_id) + (year | location/type_simple), # first DV
  
  error ~ year + partisan + log(samplesize) + days_till_election +
  (year | pollster_rating_id) + (year | location/type_simple)), # second DV
  data = d, iter = 4000
)
```

Making predictions used in the maps:

```R 
make_predictions(fit, pred = "year", 
                 at = list(location = unique(d$location), days_till_election = 0, 
                           samplesize = 800),
                 re.form = ~ (year | location), resp = "bias")
```

There's a good deal more code in the full R Markdown document used to create
this post (which can be found in my website's Github repository), but it's 
mostly just data cleaning and futzing around with Plotly to make the maps.
